The semantic drift of quotations in blogspace: a case study in short-term cultural evolution

### 2016-07-25 brains-protocol.tex
chore/todos: rework todos for results

\todo{\#16: add a quick description of what types of substitutions we see, then moving on to the observables since it's not what principally interests us.}

\todo{order of presentation:
- distance in substitution
- susceptibility of POS
- susceptibility of bins of absolute value
- variation of bins of absolute value
- susceptibility of quartiles in sentence
- variation of quartiles in sentence (bins->bins or quartile->bins or quartile->quartile)
- PCA of susceptibilities absolute, quartile, absolute+quartile
}

@@

\todo{\#19: explain if we're susceptible to bias from single-substitution}

@@

Now, for a given feature $\phi$, we obtain the mean susceptibility $\sigma_{\phi}(f)$ for the feature value $f$ by averaging this ratio over all words such that $\phi(w) = f$, that is% (only taking into account words that are substituted at least once)
:

@@

%\begin{figure*}[!th]
%    \centering
%    \includegraphics[width=\textwidth]{images/computed-figures/feature_susceptibilities.png}
%    \caption{\textbf{Substitution susceptibility:} average susceptibility to substitution versus average feature value of a candidate word for substitution, with 95\% asymptotic confidence intervals.
%    The heatmap on the lower-right shows the joint effect of Number of synonyms and Number of phonemes on susceptibility, averaged over the respective single-variable ranges, with sample size (word numbers) in parentheses.}
%    \label{fig:feature-susceptibilities}
%\end{figure*}
\medskip

\todo{\#16:
(1) show POS don't differ,
(2) note that stopwords are excluded, which excludes POS open/closed discussion
}

\todo{\#16: explain we do this on quartiles, rework results discussion}

\todo{\#16: explicit all significance claims, referring to the CIs}

\todo{\#17: show other feature variations and other models in annex}

\todo{\#15: relate to missed literature}

@@

The other features --- Age of acquisition, Number of phonemes, Clustering coefficient and Number of synonyms --- do not seem to exhibit any particularly significant effect on susceptibility.
{If we set aside the values for low Number of phonemes}, for each of these features it is indeed possible to draw a constant line which always remains within the respective confidence intervals.
If these variables have an effect, it is by no means as strong as it is for Word frequency.
This is remarkably clear for Clustering coefficient and Age of acquisition, where susceptibility values remain within quite small intervals (respectively $[.13-.18]$ and $[.16-.20]$).
{We may notice a slight effect for the lowest values of Number of synonyms and Number of phonemes, where the mean susceptibility is almost half as high as the average of the other values (respectively $.09$ vs. $.16$, and $.11$ vs. $.17$).}
Keeping in mind the poor statistical significance of this effect, we could still wonder if the shortest words and words with fewest synonyms are significantly less susceptible to substitution.
To further examine this phenomenon, we plotted the two-dimensional map of susceptibility values for these two features (see heatmap at the bottom right of Fig.~\ref{fig:feature-susceptibilities}).
%\TG{TBF: [[We see that outlier values (the two darkest cells) are both due to cells where the number of items is only slightly above the significance threshold (11 items, while we decided to cut at 10).]]}
%\TG{TBF: [[If we focus on cells based on at least several dozens of words]]}
{Even if there are a few outlier cells, values tend to navigate around the mean value ($.16$) with little obvious regularity (except for a low number of synonyms, consistent with the unidimensional graph)}. {On the whole, this makes it relatively hard to draw any conclusion as regards the direction of an effect, except for the least populated value ranges (which as a result are also less significant).}
%\TG{TBF: [[This confirms the fact that the slight late increases observed for Age of acquisition and Number of phonemes are rather non-significant.]]}

%\TB{This class of words corresponds to \todo{} [\TB{highly specialized words? domain-specific? a hard limit on too complicated words?}]}
%Age of acquisition and Number of phonemes delineate two categories of words: both features have little to no significant effect on susceptibility for low- to mid-valued words, but very high-valued words behave differently and tend to be more substituted (clearly for high Number of phonemes, and only as a tendency for high Age of acquisition).
%These high-valued words also have low occurrence frequencies, which corresponds to the slope change in the frequency curve between extreme low frequency words and mid- to high-frequency words. \rk{check this with a plot of freq vs. clustering and freq vs. AoA, or simply plot of frequency for the high-valued words for AoA and CC.}

@@

% plus ou moins choisies uniformément (modulo les réserves exhibées en 4.1), on peut montrer en quoi elles sont modifiées Finally the results for Clustering coefficient and Number of synonyms, despite some tendencies, do not allow us to conclude either to an effect or no effect of those features on substitution susceptibility.

@@

@@

We also introduce a null hypothesis $\mathcal{H}_0$ to compare the actual variation of a word's feature to its expected variation, assuming the arrival word $\warrival$ was randomly chosen from the whole pool of words available in the data set for that feature.\footnote{For instance, when considering the feature ``Clustering coefficient'', the arrival word is randomly chosen among words present in the data set of FA norms.}
%\TB{randomly chosen from the whole pool of words in the data set from which the feature is built (e.g. for Clustering coefficient: randomly chosen in the pool of words in the Free Association norms)}.\marginnote{C: ici on est bien d'accord que le pool of FA words c'est tout le set de mots, pas juste les voisins? la formulation est un peu ambigüe. S: oui et non. Oui: c'est pas juste les voisins. Non: le pool dépend de la feature, c'est pas juste FA. Il y avait une erreur, j'ai reformulé pour être plus clair.}
In this case, since $\phi(\warrival)$ becomes a constant value in the above averaging (by definition $\warrival$ does not depend on $\wstart$ anymore),  the baseline variation under $\mathcal{H}_0$ may be rewritten as:\footnote{We additionally considered an alternative null hypothesis, denoted $\mathcal{H}_{00}$, where the arrival word is randomly chosen \emph{among immediate synonyms of the start word}, that is an arrival word chosen among semantically plausible though still random words. In this case $\warrival_{00}$ does depend on $\wstart$. Our conclusions hold under this second null hypothesis, so for the sake of clarity we chose to keep the simpler~$\mathcal{H}_0$.}
$$\nu_{\phi}^0 (f) = \left<\phi\right>$$
%We thereby obtain the mean variation of feature for each start feature value, comparing the variations to a situation where arrival words are chosen randomly.

This approach yields a fine-grained view of how word features evolve upon substitution, on average, with respect to
\begin{seriate}
\item the original feature (\hbox{vs.} $y=x$) and
\item a random arrival (\hbox{vs.} $\nu_{\phi}^0$).
\end{seriate}
%\begin{figure*}[!th]
%    \centering
%    \includegraphics[width=\textwidth]{images/computed-figures/feature_variations-binned.png}
%    \caption{\textbf{Feature variation upon substitution:} $\nu_\phi$, average feature value of the appearing word as a function of the feature value of the disappearing word in a substitution, with 95\% asymptotic confidence intervals.
%    The overall position of the curve with respect to the dashed line representing $\mathcal{H}_0$ (constant $\nu_{\phi}^0$) indicates the direction of the cognitive bias.
%    The intersection with $y = x$ marks the attractor value.
%    %The fact that all curves have slopes smaller than 1 means that the substitution operation is contractile on average: each feature will converge towards its own specific asymptotic range.
%    }
%    \label{fig:feature-variations}
%\end{figure*}

@@

\todo{\#16:
(1) add H00 lines to account for semantic similarity (nothing more fancy since it involves a threshold),
(2) rework results discussion
}

\todo{\#17: show other feature variations and other models in annex}

\todo{\#17: explain binning}

\todo{\#18: tone down claims about contractile: it's a possible hypothesis if this were the only process, but not observed with the mix of all other processes}

\todo{\#15: relate to missed literature}

@@

Word frequency, with $\nu_{\phi}$ always significantly above $\nu_{\phi}^0$, exhibits a strong bias towards more frequent words. This, in turn, is consistent with the hypothesis that substitution is a recall process, since common words are favored over awkward ones, while it goes against the idea that it could be a familiarity process, where awkward terms would be favored.
%\cam{je crois qu'on peut éviter l'inférence inverse en ne disant pas "confirms" (c'est exagéré en effet) mais en disant juste que c'est cohérent avec un processus mais pas l'autre; j'ai changé le wording}
% , Mais est-ce que c'est pas une inférence inverse que de dire ceci? i.e. on sait qu'une tâche "familiarity/recognition" implique un biais en faveur des mots awkards, et qu'une tâche "recall" implique un biais en faveur des mots fréquents, mais on peut pas vraiment tirer de conclusions en utilisant la réciproque, si?}.
%However, the detailed variations of the Word frequency curve, despite an inflection for high-frequency words, do not seem to follow any significant pattern.\cam{Je serais pour supprimer cette phrase purement factuelle et pas très clear-cut non plus}

@@

\todo{\#16: add the taking-context-into-account parts, potentially moving the non-relative results discussions to here:
(1) explain and add sentence-relative susceptibility by quartiles, which has a perfectly clear interpretation + results discussion,
(2) explain and add sentence-relative variations + results discussion
(3) link those results to non sentence-relative values
}

\todo{\#17: show other feature variations and other models in annex}

\todo{\#15: relate to missed literature}

>>>>

\todo{Quick description of what types of substitutions we see (\#16)}

\todo{Mention distance in substitution}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/computed-figures/all-susceptibilities-pos.png}
    \caption{}
    \label{fig:pos-susceptibilities}
\end{figure}

@@

Now, for a given feature $\phi$, we obtain the mean susceptibility $\sigma_{\phi}(f)$ for the feature value $f$ by averaging this ratio over all words such that $\phi(w) = f$, that is:

@@

\begin{figure*}[!th]
    \centering
    \includegraphics[width=\textwidth]{images/computed-figures/paper-susceptibilities-quantilebins_global.png}
    \caption{\textbf{Substitution susceptibility:}% average susceptibility to substitution versus average feature value of a candidate word for substitution, with 95\% asymptotic confidence intervals.
    %The heatmap on the lower-right shows the joint effect of Number of synonyms and Number of phonemes on susceptibility, averaged over the respective single-variable ranges, with sample size (word numbers) in parentheses.
    }
    \label{fig:feature-susceptibilities-global}
\end{figure*}

\todo{Exemplify with POS (note stopwords excluded) (\#16)}

\todo{Discuss results for global value, then (\#15) relate to missed literature}

@@

The other features --- Age of acquisition, Number of phonemes, Clustering coefficient and Number of synonyms --- do not seem to exhibit any particularly significant effect on susceptibility.
{If we set aside the values for low Number of phonemes}, for each of these features it is indeed possible to draw a constant line which always remains within the respective confidence intervals.
If these variables have an effect, it is by no means as strong as it is for Word frequency.
This is remarkably clear for Clustering coefficient and Age of acquisition, where susceptibility values remain within quite small intervals (respectively $[.13-.18]$ and $[.16-.20]$).
{We may notice a slight effect for the lowest values of Number of synonyms and Number of phonemes, where the mean susceptibility is almost half as high as the average of the other values (respectively $.09$ vs. $.16$, and $.11$ vs. $.17$).}
Keeping in mind the poor statistical significance of this effect, we could still wonder if the shortest words and words with fewest synonyms are significantly less susceptible to substitution.
To further examine this phenomenon, we plotted the two-dimensional map of susceptibility values for these two features (see heatmap at the bottom right of Fig.~\ref{fig:feature-susceptibilities}).
{Even if there are a few outlier cells, values tend to navigate around the mean value ($.16$) with little obvious regularity (except for a low number of synonyms, consistent with the unidimensional graph)}. {On the whole, this makes it relatively hard to draw any conclusion as regards the direction of an effect, except for the least populated value ranges (which as a result are also less significant).}

@@

@@

\begin{figure*}[!th]
    \centering
    \includegraphics[width=\textwidth]{images/computed-figures/paper-variations-quantilebins_global.png}
    \caption{\textbf{Feature variation upon substitution:} %$\nu_\phi$, average feature value of the appearing word as a function of the feature value of the disappearing word in a substitution, with 95\% asymptotic confidence intervals.
    %The overall position of the curve with respect to the dashed line representing $\mathcal{H}_0$ (constant $\nu_{\phi}^0$) indicates the direction of the cognitive bias.
    %The intersection with $y = x$ marks the attractor value.
    %The fact that all curves have slopes smaller than 1 means that the substitution operation is contractile on average: each feature will converge towards its own specific asymptotic range.
    }
    \label{fig:feature-variations-global}
\end{figure*}

@@

\todo{Introduce both H0 and H00}

We also introduce a null hypothesis $\mathcal{H}_0$ to compare the actual variation of a word's feature to its expected variation, assuming the arrival word $\warrival$ was randomly chosen from the whole pool of words available in the data set for that feature.\footnote{For instance, when considering the feature ``Clustering coefficient'', the arrival word is randomly chosen among words present in the data set of FA norms.}
In this case, since $\phi(\warrival)$ becomes a constant value in the above averaging (by definition $\warrival$ does not depend on $\wstart$ anymore),  the baseline variation under $\mathcal{H}_0$ may be rewritten as:\footnote{We additionally considered an alternative null hypothesis, denoted $\mathcal{H}_{00}$, where the arrival word is randomly chosen \emph{among immediate synonyms of the start word}, that is an arrival word chosen among semantically plausible though still random words. In this case $\warrival_{00}$ does depend on $\wstart$. Our conclusions hold under this second null hypothesis, so for the sake of clarity we chose to keep the simpler~$\mathcal{H}_0$.}
$$\nu_{\phi}^0 (f) = \left<\phi\right>$$

This approach yields a fine-grained view of how word features evolve upon substitution, on average, with respect to
\begin{seriate}
\item the original feature (\hbox{vs.} $y=x$) and
\item a random arrival (\hbox{vs.} $\nu_{\phi}^0$).
\end{seriate}

@@

\todo{Update results, explaining binning and toning down contractility, and relate to missed literature}

@@

Word frequency, with $\nu_{\phi}$ always significantly above $\nu_{\phi}^0$, exhibits a strong bias towards more frequent words. This, in turn, is consistent with the hypothesis that substitution is a recall process, since common words are favored over awkward ones, while it goes against the idea that it could be a familiarity process, where awkward terms would be favored.

@@

\subsection{Sentence context}

\todo{Explain in-sentence susceptibility and sentence-relative variation, relating to global value results}

\begin{figure*}[!th]
    \centering
    \includegraphics[width=\textwidth]{images/computed-figures/paper-susceptibilities-sentencequantiles.png}
    \caption{}
    \label{fig:feature-susceptibilities-in_sentence}
\end{figure*}

\begin{figure*}[!th]
    \centering
    \includegraphics[width=\textwidth]{images/computed-figures/paper-variations-quantilebins_sentencerel.png}
    \caption{}
    \label{fig:feature-variations-sentencerel}
\end{figure*}
